Performance analysis of high-resolution \\ ice sheet simulations

### 2022-10-06 paper/perfmod.tex
make the major addition

Reader familiarity is assumed with the standard SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}.  These (continuum) models are regarded here as end members of current-usage stress-balance approximations.  Familiarity with the mass continuity and surface kinematical equations \citep{GreveBlatter2009} is also assumed.  However, before analyzing the performance consequences of numerical modeling choices, the form of the mass continuity equation will be examined, and then the terms ``explicit'' and ``implicit'' will be carefully defined.

@@

Thus, as can be addressed by linearized analysis \citep{Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of (discretized) advective and diffusive mechanisms.  Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation, for bed elevation $b(\bx)$.  A numerical thickness perturbation will often cause the velocity $\bu$ to respond by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.  In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term is along the surface gradient $\gradx s$.  One might write that Equation \eqref{eq:masscontinuity} has velocity $\bu(H,\grad_x s)$, a non-local function of geometry; a stress balance solution is required to evaluate this function.  A numerical instability of Equation \eqref{eq:masscontinuity} occurs when the ice thickness under/over-shoots its correct value because the numerical velocity from evaluating this non-local function is too strong for the given time step.

\newcommand{\nn}{\text{n}}
A diffusive description of the mass continuity equation is valid in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}:
\begin{equation}
\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a. \label{eq:siamasscontinuity}
\end{equation}
Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ in the isothermal case, where $A$ is the ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn\approx 3$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.}  While equation \eqref{eq:siamasscontinuity} does not hold directly for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed regardless.  Generically, across stress balance choices, for grounded ice sheets one observes that large values of $d$ tend to indicate locations of unstable mode growth if explicit time-steps are chosen too large.

FIXME big addition

>>>>

Reader familiarity is assumed with the standard SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}.  These (continuum) models are regarded here as end members of current-usage stress-balance approximations.  Familiarity with the mass continuity and surface kinematical equations \citep{GreveBlatter2009} is also assumed.  However, before analyzing the performance consequences of numerical modeling choices, the form of the mass continuity equation is examined, and then the terms ``explicit'' and ``implicit'' are carefully defined.

@@

Thus, as can be addressed by linearized analysis \citep{Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of (discretized) advective and diffusive mechanisms.  Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation, for bed elevation $b(\bx)$.  A numerical thickness perturbation will often cause the velocity $\bu$ to respond by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.  In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term effectively acts along the surface gradient $\gradx s$ because of the ice-domain geometry.  One might write that Equation \eqref{eq:masscontinuity} has velocity $\bu(H,\gradx s)$, a non-local function of geometry; a stress balance solution is required to evaluate this function.  A numerical instability of Equation \eqref{eq:masscontinuity} occurs when the ice thickness under/over-shoots its correct value because the numerical velocity from evaluating this non-local function is too strong for the given time step.

\newcommand{\nn}{\text{n}}
A diffusive description of the mass continuity equation is also valid in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}:
\begin{equation}
\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a. \label{eq:siamasscontinuity}
\end{equation}
Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity.\footnote{In detail, $C = 2 A (\rho g)^\nn/(\nn+2)$ in the isothermal case, where $A$ is the ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn\approx 3$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.}  While Equation \eqref{eq:siamasscontinuity} does not hold directly for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed regardless.  Generically across stress balance choices, for grounded glaciers and ice sheets one observes that large values of $d$ indicate locations of unstable mode growth if explicit time-steps are chosen too large.

However, either statement \eqref{eq:masscontinuity} or \eqref{eq:siamasscontinuity} of the mass continuity equation is fundamentally incomplete without modeling the changing (map-plane) boundary position of the glacier or ice sheet.  The problem of determining the updated margin position is only well-posed via an inequality constraint, namely that ice thickness is nonnegative ($H\ge 0$), equivalently that the ice surface elevation equals or exceeds the bed elevation ($s \ge b$).  The mathematical role of this constraint has been understood for some time in the SIA model \citep{Calvoetal2002,JouvetBueler2012,SchoofHewitt2013}, but its role as the source of the margin location is universal across fluid-layer problems with signed source terms \citep{Bueler2021conservation}.

It is in this context that we can distinguish time-stepping types in coupled geometry-velocity, i.e.~mass and momentum conserving, glacier and ice sheet models.  A coupled model computes (\emph{fully}) \emph{implicit} time steps if no significant aspects of the geometry or velocity are held fixed during the (coupled) solution.  That is, in an implicit scheme a coupled and well-posed model for the updated ice surface elevation (or thickness) and velocity fields is solved.  A scheme which does not allow the map-plane ice-covered region to change during a time step cannot respond physically (conservatively) to a change in climate inputs; such a scheme is not actually implicit for the coupled problem and it cannot validly compute long-duration time steps.  Thus we say that a coupled scheme is \emph{semi-implicit} if, in particular, aspects of the ice geometry are held fixed during its coupled geometry-velocity solution.  A scheme is only semi-implicit if it computes margin advance or retreat after the velocity field is updated.  Finally, a scheme is \emph{explicit} for the coupled problem, and uses a time-step \emph{splitting} scheme \citep{HundsdorferVerwer2003}, if the ice geometry is held fixed during the velocity solution, and then this (accepted) velocity solution is used to update the geometry, e.g.~through an advection step for Equation \eqref{eq:masscontinuity}.

Thus a key point about time-stepping schemes for the coupled geometry and velocity problem for glaciers and ice sheets is that implicit schemes are not merely solving fixed-domain systems of coupled PDEs.  The mathematical problem for the coupled and implicit time-step includes the condition which controls the free boundary, namely the inequality constraint of nonnegative thickness \citep{Bueler2021conservation}.  Implicit schemes which solve these free-boundary problems at each time step can, at high spatial resolution, transcend stability limitations on time step duration.  They can do time-stepping at a rate controlled by climatic-input time scales.
